\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-21 doc/paper-jss/sdmTMB-paper.Rnw
Updates

@@

\Abstract{Geostatistical data are common across scientific fields and available
  in increasingly large datasets. However, appropriate models to analyse these
  data, such as generalised linear mixed effects models with Gaussian random
  fields, are computationally intensive and challenging to implement for many
  users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends
  the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or
  \pkg{mgcv} to include spatial and spatiotemporal latent fields using
  a predictive-process SPDE-(stochastic partial differential equation) based
  approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and
  estimation is conducted via maximum marginal likelihood with Template Model
  Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and
  \pkg{rstan}. We describe the model and explore case studies that illustrate
  \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary
  processes (time-varying and spatially varying coefficients), and anisotropy
  (directionally dependent spatial correlation). Additional functionality
  includes hurdle models, break-point effects, and out-of-sample cross
  validation. Finally, we compare the functionality, speed, and interfaces of
  related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude
  faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible
interface will help open this useful class of models to a wider field of
geostatistical analysts.}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\widowpenalty10000 \clubpenalty10000

@@

A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
variables that are not otherwise accounted for in the model. Alternatively,
$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some
function that defines the rate that spatial covariance decays with
distance,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
\]
Various covariance functions are possible, but a popular and flexible one is the
Mat\'ern \citep{matern1960} (Figure~\ref{fig:matern-range}). If we define $\lVert h \rVert$ as the Euclidean
distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
Gamma function, $K_\nu$ represents the modified Bessel function of the second
kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
$\nu$ controls the smoothness of the covariance function, but can be fixed at
$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
as $\sqrt{8} / \kappa$ and (2) marginal standard deviation as
$1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
The range is the distance that correlation decays to $\approx 0.13$ (Figure~\ref{fig:matern-range}).

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="TODO">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("Range = 0.2") +
  labs(x = "X", y = "Y")

s2 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.6,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("Range = 0.6") +
  labs(x = "X", y = "Y")

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = c(0.7, 0.7)) +
  ggtitle("Matérn correlation function")

cowplot::plot_grid(sim_g3, sim_g1, sim_g2, align = "h", ncol = 3L)
@

The assumption that correlation decays equally in all directions can be relaxed
to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated
parameters governing the major axis direction of geometric anisotropy and the
ratio of the major and minor axes \citep{lindgren2011, thorson2015ices,
thorson2018comp}.

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
expensive and methods of working directly with the sparse precision matrix
($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
the SPDE approach to working with the precision matrix in section \ref{sec:software},
but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim
  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian
random fields for each time step $t$, $\bm{\epsilon}_{t}$
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]
In this case we assume the spatiotemporal random fields are independent at each
time step, but we could alternatively assume they are structured as a random
walk or autoregressive process (demonstrated in examples below). The
spatiotemporal random fields represent latent variables causing spatial
correlation that changes with each time step.

@@

In practice, the above models can become considerably more complex by, for
example, letting coefficients vary though time, letting coefficients vary
through space \citep{hastie1993} as random fields, or adding random intercepts
by grouping factors. All these components are additive in link space. Adopting
the notation ``main'' for main effects, ``tvc'' for time-varying coefficients,
and ``svc'' for spatially varying coefficients, a maximally complex model could
be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector
of coefficients that are allowed to vary through time as random walks or AR(1)
processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following
a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim
\operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an
offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and
$\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal
intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate
these model components in the examples later in this paper.

@@

So far we have described models with one linear predictor and one family (common
terminology in \proglang{R} packages for the combination of an observation
likelihood and link). Frequently, data are better represented with two-part
``delta'' or ``hurdle'' models, which include linear predictors and observation
distributions for two processes: zero vs.\ non-zero values and positive values
\citep{aitchison1955}. Here, we describe two types of delta models. We drop the
space and time subscripts for simplicity and use $\bm{X_1} \bm{\beta_1}$ and
$\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors
for the first and second model components.

@@

Using $p$ to denote the probability of a non-zero observation and $r$ to
represent the expected rate for positive data, we can construct linear
predictors in link space
\[
\begin{aligned}
\operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
\log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\end{aligned}
\]
We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood
distribution (e.g., lognormal or gamma with $\phi$ representing a generic
dispersion parameter). We use $\mathrm{I}(.)$ to denote an indicator function,
which is $1$ if $y > 0$ and 0 if $y = 0$.
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

The expectation for a new data point $y$ is then $p \cdot r$.

@@

An additional delta-model is possible that has several advantages over the
logit-log delta model \citep{thorson2018poisson}. The primary advantage is that
both linear components use a log link so the effects are additive in log space
and can be combined. There are still two linear predictors, but they represent
a theoretical group number $n$ and a theoretical weight per group $w$
\citep{thorson2018poisson}.
\[
\begin{aligned}
\log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
\log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link and
the linear predictor predicts group number $n$, \emph{not} positive observation
probability $p$. These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link
\citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but
the group number $n$ also enters into the expected positive rate $r$. These
probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and
positive observation distribution as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

@@

% \begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
% \end{landscape}

@@

We find that the Poisson-link delta-lognormal model is favoured by AIC. In an
applied situation we would consider comparing models with cross validation and
inspecting the distribution of residuals.

\subsection{Adding AR(1) random fields and comparing isotropic correlation}

@@

The AR(1) fields can be represented as:

>>>>

\newcommand{\sdmTMB}{\pkg{sdmTMB}}
\newcommand{\INLA}{\proglang{R}-\pkg{INLA}}
\newcommand{\TMB}{\pkg{TMB}}
\newcommand{\s}{\bm{s}}
\newcommand{\R}{\proglang{R}}

@@

\Abstract{
Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

@@

@@

% \widowpenalty10000
% \clubpenalty10000

@@

A basic GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$. This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model. Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise. More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}ab).

The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some function that defines the rate at which spatial covariance decays with distance,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
\]
Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}). If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate. The parameter $\nu$ controls the smoothness of the covariance function. In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}. A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent. A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right)
\]

% Assuming $\nu = 1$, the parameters $\kappa$ and $\tau$ can be converted to two useful values: correlation range ($\sqrt{8} / \kappa$) and marginal standard deviation The range is the distance that correlation decays to $\approx 0.13$ .

% ($1 / \sqrt{4 \pi \tau^2 \kappa^2} $).

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="Example Gaussian random fields for two range values. The range describes the distance at which spatial correlation decays to $\\approx 0.13$ in coordinate units (i.e., the distance at which two points are effectively independent). Panel a shows a shorter range than panel b, which results in a ``wigglier'' surface. Panel c shows the Mat\\'ern function for these two range values. The dashed horizontal line shows a correlation of 0.13.">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(a) Range = 0.2") +
  labs(x = "X", y = "Y")

s2 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.6,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("(b) Range = 0.6") +
  labs(x = "X", y = "Y")

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = c(0.7, 0.7)) +
  ggtitle("(c) Matérn correlation function")

cowplot::plot_grid(sim_g1, sim_g2, sim_g3, align = "h", ncol = 3L)
@

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive and methods for working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are more efficient. We introduce the SPDE approach to working with the precision matrix in Section \ref{sec:software}, but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim
  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]
In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}). The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors. All these components are additive in link space. Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate these model components in the examples below.

@@

So far we have described models with one linear predictor and one family (common terminology in \proglang{R} packages for the combination of an observation likelihood and link). Frequently, data are better represented with two-part ``delta'' or ``hurdle'' models, which include linear predictors and observation distributions for two processes: zero vs.\ non-zero values and positive values, respectively \citep{aitchison1955}. Here, we describe two types of delta models, dropping the space and time subscripts for simplicity.
% and use $\bm{X_1} \bm{\beta_1}$ and $\bm{X_2} \bm{\beta_2}$ to refer to generic model matrices and parameter vectors for the first and second model components.

@@

Using $p$ to denote the probability of a non-zero observation and $r$ to represent the expected rate for positive data, we can construct two linear predictors ($\eta_1$ and $\eta_2$) in link space
\[
\begin{aligned}
% \operatorname{logit} (p) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (r) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\operatorname{logit} (p) &= \eta_1,\\
\log (r) &= \eta_2.
\end{aligned}
\]
We can relate $p$ and $r$ to the data via a Bernoulli and positive likelihood distribution (e.g., lognormal or gamma with $\phi$ representing a generic dispersion parameter). We use $\mathrm{I}(.)$ to denote an indicator function, which is $1$ if $y > 0$ and 0 if $y = 0$.
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

The expectation for a new data point is then the probability of a non-zero multiplied by the positive rate: $p \cdot r$.

@@

An additional delta-model is possible that has several advantages over the logit-log delta model \citep{thorson2018poisson}. The primary advantage is that both linear predictors use a log link so the linear predictors can be added in link space. In these models, the linear predictors represent a theoretical group number $n$ and a theoretical weight (e.g., mass) per group $w$ \citep{thorson2018poisson},
\[
\begin{aligned}
% \log (n) &= \boldsymbol{X_1} \boldsymbol{\beta_1} + \ldots,\\
% \log (w) &= \boldsymbol{X_2} \boldsymbol{\beta_2} + \ldots.
\log (n) &= \eta_1,\\
\log (w) &= \eta_2.
\end{aligned}
\]
Note that the first linear predictor has a log link, \emph{not} a logit link, and the linear predictor predicts group number $n$, \emph{not} positive observation probability $p$. These get transformed \citep{thorson2018poisson} via
\[
\begin{aligned}
p &= 1 - \exp(- \exp( \log n)),\\
r &= \frac{n w}{p}.
\end{aligned}
\]
The first part is the complementary-log-log inverse link \citep[][p.~31]{mccullagh1989} (which has its roots in a Poisson process) but the group number $n$ also enters into the expected positive rate $r$. These probabilities $p$ and positive rates $r$ are then entered into a Bernoulli and positive observation likelihood as before
\[
\begin{aligned}
\mathrm{I}(y > 0) &\sim \operatorname{Bernoulli} \left( p \right),\\
y | y > 0 &\sim \operatorname{Positive-distribution} \left( r, \phi \right).
\end{aligned}
\]

@@

The design goal of \sdmTMB\ was to make...

\begin{figure}[p]
\centering
\includegraphics[width=4in]{flowchart}
\caption[]{
TODO
}\label{fig:flowchart}
\end{figure}

\sdmTMB\ relies on several well-established \R\ packages (FIG?).
\sdmTMB\ uses \pkg{fmesher} and \INLA\ to generate input matrices for SPDE calculations \citep[$\bm{C}$, $\bm{G}$, and $\bm{K}$ in][]{lindgren2011}.
The SPDE calculations themselves are performed by functions within \pkg{TMB} (\texttt{R\_inla} namespace).
Data structures required to implement penalized smoothers are constructed with \pkg{mgcv}'s \texttt{smooth2random()} function similarly to \pkg{gamm4} and \pkg{brms}.
The statistical model itself is written as a \proglang{C++} template for \TMB.
\TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \proglang{R} with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior distribution for Bayesian inference.

Is modular in nature...

@@

% \begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples,
notation, and example code. An \pkg{sdmTMB} model can be built from any
combination of the process components (first six rows) plus an observation
component (last row). The examples are from an SDM context, but the model can be
fit to any spatially referenced point data. Notation: We refer to design
matrices as $\bm{X}$. The indexes $\bm{s}$, $t$, and $g$ index spatial
coordinates, time, and group, respectively. The $\sigma$ and $\bm{\Sigma}$
symbols represent standard deviations and covariance matrices, respectively. All
other symbols refer to the  described model components (e.g., $\bm{\beta}$ and
$\bm{\omega}$ refer to a vector of main effects and spatial random field
deviations, respectively). See the \pkg{sdmTMB}
\href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette}
for a full description of the model. Note that \code{s()} denotes a smoother as
in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint
`hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes
a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in
an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
% \end{landscape}

@@

We find that the Poisson-link delta-lognormal model is favoured by AIC. In an applied situation we would inspect the distribution of the residuals and consider comparing models with cross validation.

\subsection{Adding AR(1) random fields and comparing isotropic correlation}\label{sec:dogar1}

@@

The AR(1) fields can be represented as:
\[
\begin{aligned}
\bm{\epsilon}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\epsilon}_{t>1} &= \rho \bm{\epsilon}_{t-1} + \sqrt{1 - \rho^2} \bm{\delta}_{t},  \:
\bm{\delta}_{t} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where $\rho$ represents the estimated autoregressive parameter allowing the spatial field at time $t$ to be correlated with the spatial field at time $t-1$ with deviations created by $\bm{\delta}_t$, which are themselves independent random fields by year.
